---
title: "cyclistic bike share case study"
output: html_document
date: '2022-07-20'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Information about company
A bike-sharing company called Cyclistic wants to analyze their user data to find the differences in user behaviour between their two types of users which are the casual, who pay per ride and the annual member who pays a yearly subscription for the service. From this analysis we will gather insights that the marketing team will use to design a new marketing strategy to convert casual riders into annual members.

Load all libraries.

```{r}
library(tidyverse) 
library(lubridate)  
library(ggplot2)  
library(dplyr)
library(ggthemes)
```

Load all the data.

```{r}
April_2020 <- read.csv("../input/cyclistic/202004-divvy-tripdata/202004-divvy-tripdata.csv")
May_2020 <- read.csv("../input/cyclistic/202005-divvy-tripdata/202005-divvy-tripdata.csv")
June_2020 <- read.csv("../input/cyclistic/202006-divvy-tripdata/202006-divvy-tripdata.csv")
July_2020 <- read.csv("../input/cyclistic/202007-divvy-tripdata/202007-divvy-tripdata.csv")
August_2020 <- read.csv("../input/cyclistic/202008-divvy-tripdata/202008-divvy-tripdata.csv")
September_2020 <- read.csv("../input/cyclistic/202009-divvy-tripdata/202009-divvy-tripdata.csv")
October_2020 <- read.csv("../input/cyclistic/202010-divvy-tripdata/202010-divvy-tripdata.csv")
November_2020 <- read.csv("../input/cyclistic/202011-divvy-tripdata/202011-divvy-tripdata.csv")
December_2020 <- read.csv("../input/cyclistic/202012-divvy-tripdata/202012-divvy-tripdata.csv")
January_2021 <- read.csv("../input/cyclistic/202101-divvy-tripdata/202101-divvy-tripdata.csv")
February_2021 <- read.csv("../input/cyclistic/202102-divvy-tripdata/202102-divvy-tripdata.csv")
March_2021 <- read.csv("../input/cyclistic/202103-divvy-tripdata/202103-divvy-tripdata.csv")
```

## Analysis
Check the column names.

```{r}
colnames(April_2020)
colnames(May_2020)
colnames(June_2020)
colnames(July_2020)
colnames(August_2020)
colnames(September_2020)
colnames(October_2020)
colnames(November_2020)
colnames(December_2020)
colnames(January_2021)
colnames(February_2021)
colnames(March_2021)
```

Check the structure of the data.

```{r}
str(April_2020)
str(May_2020)
str(June_2020)
str(July_2020)
str(August_2020)
str(September_2020)
str(October_2020)
str(November_2020)
str(December_2020)
str(January_2021)
str(February_2021)
str(March_2021)
```

Mutate the 2020 data to match the 2021 data.

```{r}
April_2020 <-  mutate(April_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 
May_2020 <-  mutate(May_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))
June_2020 <-  mutate(June_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))
July_2020 <-  mutate(July_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))
August_2020 <-  mutate(August_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))
September_2020 <-  mutate(September_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id)) 
October_2020 <-  mutate(October_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))
November_2020 <-  mutate(November_2020, start_station_id = as.character(start_station_id)
                   ,end_station_id = as.character(end_station_id))
```

Join all the data from 2020 and 2021 together.

```{r}
all_trips <- bind_rows(April_2020, May_2020, June_2020, July_2020, August_2020, September_2020, October_2020, November_2020, December_2020, January_2021, February_2021, March_2021)
```

Remove start_lat, start_lng, end_lat and end_lng from the data.

```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

Check the columns, number of rows, dimensions, head, structure and summary of the data.

```{r}
colnames(all_trips)  
nrow(all_trips)  
dim(all_trips)  
head(all_trips)  
str(all_trips)  
summary(all_trips)
```

Recode the data.

```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```

Create a categorical representation of the data by showing the variable name and frequency in the table.

```{r}
table(all_trips$member_casual)
```

Format the data and seperate the dates into month, day, year and day of the week.

```{r}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

Calculate the time difference between the two objects.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

Check the structure of the data.

```{r}
str(all_trips)
```

Convert the ride_length from factor to numeric so calculations can be run on the data.

```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Create a new version of the dataframe (v2).

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

Check structure of the new data (v2).

```{r}
str(all_trips_v2)
```

Convert ride_length from factor to numeric.

```{r}
is.factor(all_trips_v2$ride_length)
all_trips_v2$ride_length <- as.numeric(as.character(all_trips_v2$ride_length))
is.numeric(all_trips_v2$ride_length)
```

Analysis on ride_length with the figures in seconds.

```{r}
mean(all_trips_v2$ride_length) 
median(all_trips_v2$ride_length) 
max(all_trips_v2$ride_length) 
min(all_trips_v2$ride_length)
```

Use summary() on the data.

```{r}
summary(all_trips_v2$ride_length)
```

Compare the members and casual users.

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

Check the average ride time by each day for the members vs casual users.

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

The days of the week are not in order, that needs to get fixed ordering them correctly.

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Check the average ride time by each day for the members vs casual users.

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

Do analysis on the ridership data by type and weekday.

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n() 
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)
```

## Visualizations
Visualize the number of rides by the rider type.

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) +
  geom_col(alpha = 0.7, position = "dodge") +
  theme_fivethirtyeight() +
  scale_size(range=c(2,12), guide="none") +
labs(title = "Number of Rides by User",
     x = "Day of the week",
     y = "Number of rides",
     fill = "Member vs Casual",
     caption = "Source: Google Capstone") +
theme(axis.title = element_text(),
    text = element_text(family = "Amiri"),
    legend.text=element_text(size=10)) +
  scale_fill_brewer(palette = "Set2")
```

Create a visualization for average duration.

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(alpha = 0.7, position = "dodge") +
  theme_fivethirtyeight() +
  scale_size(range=c(2,12), guide="none") +
labs(title = "Average ride duration by weekday",
     x = "Day of the week",
     y = "Average duration of rides",
     fill = "Member vs Casual",
     caption = "Source: Google Capstone") +
theme(axis.title = element_text(),
    text = element_text(family = "Amiri"),
    legend.text=element_text(size=10)) +
  scale_fill_brewer(palette = "Set2")
```

Type of bike used by the riders.

```{r}
all_trips_v2 %>% 
  group_by(rideable_type, member_casual) %>%
  summarise(number_of_ride = n()) %>%
  ggplot(aes(x= rideable_type, y= number_of_ride, fill= member_casual)) +
  geom_col(alpha = 0.7, position = "dodge") +
  theme_fivethirtyeight() +
  scale_size(range=c(2,12), guide="none") +
labs(title = "Type of bike used by the users",
     x = "Type of bike",
     y = "Number of rides",
     fill = "Member vs Casual",
     caption = "Source: Google Capstone") +
theme(axis.title = element_text(),
    text = element_text(family = "Amiri"),
    legend.text=element_text(size=10)) +
  scale_fill_brewer(palette = "Set2")
```

## Findings
After conducting analysis on the data that was given, these are the findings:

- Casual riders ride more during the weekend on Saturday and Sunday.

- Member riders use the bikes consistently during the weekday.

- The preferred bike type for both the members and casuals is the docked_bike.

## Conclusions:

The marketing team should try and incorporate a reward system for longer rides since casual riders ride more on average. They should also introduce and initiate a members promotion for weekends to encourage members to ride more during weekends. Also they need to initiate weekday promotional rates for casual riders that will encourage them to becomes full members.

Thank you for going trough the analysis!



